\documentclass[10pt]{article}

\usepackage[letterpaper, portrait, margin=1.25in]{geometry}

\usepackage{authblk}
\usepackage[yyyymmdd,hhmmss]{datetime}
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{ASTR 511: Galaxies as Galaxies}
\rhead{Winter 2017}
\rfoot{\em \tiny Compiled on \today\ at \currenttime}
\cfoot{\thepage}
\lfoot{}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% author-defined commands
\newcommand\about     {\hbox{$\sim$}}
\newcommand\x         {\hbox{$\times$}}
\newcommand\othername {\hbox{$\dots$}}
\def\eq#1{\begin{equation} #1 \end{equation}}
\def\eqarray#1{\begin{eqnarray} #1 \end{eqnarray}}
\def\eqarraylet#1{\begin{mathletters}\begin{eqnarray} #1 %
                  \end{eqnarray}\end{mathletters}}
\def\non    {\nonumber \\}
\def\DS     {\displaystyle}
\def\E#1{\hbox{$10^{#1}$}}
\def\sub#1{_{\rm #1}}
\def\case#1/#2{\hbox{$\frac{#1}{#2}$}}
\def\about  {\hbox{$\sim$}}
\def\x      {\hbox{$\times$}}
\def\ug               {\hbox{$u-g$}}
\def\gr               {\hbox{$g-r$}}
\def\ri               {\hbox{$r-i$}}
\def\iz               {\hbox{$i-z$}}
\def\a                {\hbox{$a^*$}}
\def\O                {\hbox{$O$}}
\def\E                {\hbox{$E$}}
\def\Oa               {\hbox{$O_a$}}
\def\Ea               {\hbox{$E_a$}}
\def\Jg               {\hbox{$J_g$}}
\def\Fg               {\hbox{$F_g$}}
\def\J                {\hbox{$J$}}
\def\F                {\hbox{$F$}}
\def\N                {\hbox{$N$}}
\def\dd               {\hbox{deg/day}}
\def\mic              {\hbox{$\mu{\rm m}$}}
\def\Mo{\hbox{$M_{\odot}$}}
\def\Lo{\hbox{$L_{\odot}$}}
\def\comm#1           {\tt #1}
\def\refto#1          {\ref #1}
\def\T#1              {({\bf #1})}
\def\H#1              {({\it #1})}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\title{ASTR 511: Galaxies as Galaxies}
\author{Mario Juri\'{c} \& \v{Z}eljko Ivezi\'{c}}
\affil{University of Washington, Winter Quarter 2017}
\date{\vspace{-5ex}}

\begin{document}
\maketitle

\vskip 0.3in
\leftline{{\bf  Location and Time:} Tuesdays and Thursdays 2:20-3:40, PAB B305A}

\vskip 0.2in
\leftline{{\bf  Office Hours:} { Any time when our office doors are open; }}
\leftline{\hskip 1.15in After class (MJ) or Tue and Wed mornings (ZI) are the best.}
\vskip 0.2in
\leftline{{\bf  Grading:} closed-book final exam 20\%, 3 homeworks 60\%, term
                          paper presentation 20\%;}
\leftline{                \hskip 0.8in key: $>$90\%=A, $>$80\%=B, $>$70\%=C, $>$50\%=D.}
\vskip 0.2in
\leftline{{\bf  Class web site:} http://research.majuric.org/public/teaching/astr511/}
\vskip 0.2in
\leftline{{\bf  Required reading:} Ivezi\'{c}, Beers \& Juri\'{c} 2012, ARA\&A, 50, 251.}
\vskip 0.2in
\leftline{{\bf  Reference Books:} Binney \& Merrifield: {\it Galactic Astronomy}}    
\leftline{   \hskip 1.44in         Reid \& Hawley: {\it New Light on Dark Stars}} 
\leftline{   \hskip 1.44in         Binney \& Tremaine: {\it Galactic Dynamics}}
\leftline{   \hskip 1.44in         Sparke \& Gallagher: {\it Galaxies in the Universe}}
\vskip 0.3in

\leftline{\bf  The main Goals for this Class are:}
\begin{enumerate}
  \item Familiarizing students with the motivation for studying galaxies in general, and the Milky 
            Way in particular. Overview of the most relevant literature. Overview of 
            galaxy formation theories, galaxy dynamics, and connection to dark matter.
  \item Informing students about the current research in galactic astronomy, including discussion of 
            the expected observational and theoretical progress for the next decade. 
  \item Exposing students to practical problems through hands-on seminars, with emphasis
            on numerical methods and data mining, and using modern software engineering tools. 
  \item Training students to give professional talks. 
\end{enumerate} 

\vskip 0.2in
\leftline{\bf  Class Schedule (very tentative and subject to change)}
\begin{enumerate}
  \item {\bf Tue: Jan 3}  {\bf NO CLASS}
  \item {\bf Thu: Jan 5}   Introduction to course  (MJ \& ZI) 
%                              - Introduce the syllabus
%                              - Class mailing list
%                              - Have them get github accounts
%                              - Github student developer pack (https://education.github.com/)
%                              - Discuss problem sets
%                                - Mention Anaconda for Python
  \item {\bf Tue: Jan 10}  L1: Review of stellar astrophysics (ZI)
%                              - basic stellar parameters
%                              - open and globular clusters
%                              - simple stellar populations and population synthesis
  \item {\bf Thu: Jan 12}  L2: Basic properties of the Milky Way and the Local Group (ZI)
%                              - the main structural components of the MW
%                              - stars, gas, dust, black hole
%                              - brief overview of metallicity and kinematics distributions
%                              - LMC/SMC
%                              - Sgr dwarf, tidal streams, dwarf satellites
%                              - multi-wavelength view of the Milky Way
  \item {\bf Tue: Jan 17}  L3: Basic properties of galaxies (ZI)
%                              - brief overview of history
%                              - morphological differences and Hubble tuning fork diagram
%                              - exponential vs. deVac intensity profiles
%                              - multi-wavelength views of galaxies
%                              - origins of infrared emission in galaxies
%                              - starburst galaxies and radio galaxies
 \item {\bf Thu: Jan 19}  L4: Luminosity and mass functions of galaxies (ZI)
%                              - color-luminosity distributions
%                              - the spatial distribution of galaxies (vs. galaxy properties)
%                              - Butch-Oemler effect (density-morphology relations)
%                              - methods for estimating luminosity and mass functions
%                              - introduction to galaxy formation theories
%                              - galaxy mergers
%                              - dwarf galaxies
  \item {\bf Tue: Jan 24}  L5: Stellar kinematics in galaxies (MJ)
%                              - introduction to proper motion and radial velocity measurements
%                              - empirical Tully-Fisher and Faber-Jackson relations
  \item {\bf Thu: Jan 26}  L6: Galactic Dynamics I: Potentials and Orbits (MJ) {\bf HW 1 due}
%                              - potential theory
%                              - the orbits of stars
%                              - the collisionless Boltzman equation
  \item {\bf Tue: Jan 31}  L7: Galactic Dynamics II: Galaxy models (MJ)
%                              - the Jeans equations
%                              - the virial theorem
%                              - King models
  \item {\bf Thu: Feb 2}   L8: Galactic Dynamics III: Advanced Dynamics (MJ)
%                              - action-angle variables
%                              - distribution functions
  \item {\bf Tue: Feb 7}   L9: Galactic Dynamics IV: Instabilities and Resonances (MJ)
%                              - non-equilibrium behaviors of many-body systems
  \item {\bf Thu: Feb 9}   L10: Evidence for dark matter in galaxies (MJ)
%                              - rotation curves
%                              - velocity dispersion
%                              - weak lensing
%                              - MOND as an alternative hypothesis
  \item {\bf Tue: Feb 14}  L11: Discussion of homeworks and term papers (MJ \& ZI) {\bf HW 2 due}
  \item {\bf Thu: Feb 16}  L12: Stellar count distribution in the Milky Way (MJ)
%                              - derive analytic Euclidean counts
%                              - review history of counting stars (Bahcall & Soneira, etc)
%                              - modern SDSS and 2MASS results
  \item {\bf Tue: Feb 21}  L13: Stellar metallicity and kinematics in the Milky Way (ZI)
%                              - [Fe/H] and [alpha/Fe] as probes of chemical evolution
%                              - SDSS, RAVE and other modern results
%                              - SDSS results for kinematics of stars in the Milky Way
%                              - SDSS constraints for dark matter in the Milky Way disk and halo
  \item {\bf Thu: Feb 23}  L14: Open questions in galactic astronomy and modern sky surveys (ZI)
%                              - top open questions in galaxy astrophysics
%                              - Gaia, LSST, RAVE, SDSS: science drivers, data products, and data release cadence
  \item {\bf Tue: Feb 28}  Term paper presentations
  \item {\bf Thu: Mar  2}  Term paper presentations {\bf HW 3 due}
  \item {\bf Tue: Mar  7}  {\bf NO CLASS}
  \item {\bf Thu: Mar 9}   FINAL EXAM
\end{enumerate}


\vskip 0.2in

\leftline{\bf  Homework}

There will be three homeworks, designed as term projects. All three will be similar
from the technical point of view: they will involve reading a number of vectors with several 
million elements from provided files (using python and ipython notebooks), simple operations 
with these vectors such as binning and low-order statistics, and visualization of your results. 

The problems and links to data files will be posted at the class website about two weeks 
before the homework is due. They will address these three general themes:

\begin{enumerate}
\item Stellar number density distribution in MW
\item Determination of luminosity function 
\item Stellar metallicity distribution in MW  
\end{enumerate}

We will use modern software engineering tools, such as github for version control and ipython 
notebooks for HW submission. 

\vskip 0.2in

\leftline{\bf  Selected Papers for Presentation:} Pick one by Jan 22, and send
us e-mail with your choice.

\begin{enumerate}
\item Bahcall, J.N. \& Tremaine, S. 1981 (ApJ 244, 805)
       {\it Methods for determining the masses of spherical systems. I - Test particles around a point mass}
\item Dehnen, W. \& Binney, J.J. 1998 (MNRAS 298, 387)
          {\it Local stellar kinematics from HIPPARCOS data}
\item Helmi, A. \& White, S.D.M. 2001 (MNRAS 323, 529)
         {\it Simple dynamical models of the Sagittarius dwarf galaxy}
\item Ibata, R., et al. 2001 (ApJ 551, 294)
         {\it Great Circle Tidal Streams: Evidence for a Nearly Spherical Massive Dark Halo around the Milky Way}
\item Jaffe, W. 1983 (MNRAS 202, 995) 
          {\it A simple model for the distribution of light in spherical galaxies}
\item Kuijken, K. \& Gilmore, G. 1991 (ApJ 367, L9)  
     {\it The galactic disk surface mass density and the Galactic force K(z) at Z = 1.1 kpc}
\item Kuijken, K. \& Tremaine, S. 1994 (ApJ 421, 178)
            {\it On the ellipticity of the Galactic disk}
\item Navarro, J.F., Frenk, C.S. \& White, S.D.M. 1996 (ApJ 462, 563) 
         {\it The Structure of Cold Dark Matter Halos}              
\item Spitzer, L. \& Schwarzschild, M. 1953 (ApJ 118, 106) 
           {\it The Possible Influence of Interstellar Clouds on Stellar Velocities. II}
\item van den Bosch, F.C. \& Dalcanton, J.J. 2000 (ApJ 534, 146)
         {\it Semianalytical Models for the Formation of 
               Disk Galaxies. II. Dark Matter versus Modified Newtonian Dynamics}
\item Walsh, Willman \& Jerjen 2009 (AJ, 137, 450) 
   {\it The Invisibles: A Detection Algorithm to Trace the Faintest Milky Way Satellites}  
\item Johnston et al. 2008 (ApJ, 689, 936)
    {\it Tracing Galaxy Formation with Stellar Halos}
\item  Kauffmann et al. 2003 (MNRAS, 341, 33) 
   {\it Stellar Masses and Star Formation Histories for 80,000 Galaxies from the Sloan Digital Sky Survey} 
\item McGurk et al. 2010 (AJ, 139, 1261) 
     {\it Principal Component Analysis of SDSS Stellar Spectra}
\item Sesar et al. 2010 (ApJ, 708, 717)
  {\it Light Curve Templates and Galactic Distribution of RR Lyrae Stars from Sloan Digital Sky Survey Stripe 82}
\item Schlaufman et al. 2009 (ApJ, 703, 2177)
  {\it Insight into the Formation of the Milky Way Through Cold Halo Substructure. I. The ECHOS of Milky Way Formation}
\item Law and Majewski 2010 (ApJ, 714, 229) 
 {\it The Sagittarius Dwarf Galaxy: A Model for Evolution in a Triaxial Milky Way Halo}
\item Beers et al. 2012 (ApJ, 746, 34) 
 {\it The Case for the Dual Halo of the Milky Way}
\item Berry et al. 2012 (ApJ, 757, 166)
 {\it The Milky Way Tomography with SDSS. IV. Dissecting Dust}
\item Evans \& Williams 2014 (MNRAS, 443, 791)
 {\it A very simple cusped halo model} 
\end{enumerate}


These papers are very relevant for our class and cover subjects that are
not going to be discussed in detail. Choose a paper and prepare a 10 min long 
Powerpoint (or equivalent) presentation
(presumably including the most important figures from the paper). Pretend 
you did the work yourself and are giving an invited talk at a meeting. 
There will be a 5 min long question and answer session after the talk. 
The purpose of this exercise is to 1) learn some science, 2) practice 
extracting relevant information from papers 3) practice giving talks,
4) practice having over-active audience.

Do not forget the following good practices: 1) empty your pockets (no loose 
change, keys, phone, and such),
2) talk slowly and sufficently loud, 3) don't look at the floor, control
your audience with direct eye contact, don't turn your back to the audience, 
don't be agressive with the pointer, etc. 4) don't rush (don't overload 
your 10 min long presentation), concentrate on the most important points,
5) emphasize what are truly new, and, possibly, unexpected results.
6) comment on the limitations and pitfalls of the presented analysis;
how do we know it's right, could it be wrong?


\end{document}
